---
title: "Validate DIO Model"
output:
  md_document:
    variant: gfm
---

## Load the RDS of the built model
```{r setup, include=TRUE}
# Load useeior to get access to built-in validation functions.
library(useeior)
packageVersion("useeior")

# Read the built DIO v2.0 model
DIO <- readRDS("model/DIO.rds")


```
## Basic Model Info
```{r}
print(paste("The model name is", DIO$specs$Model))

print(paste("The model has",length(DIO$Commodities$Code), "commodities,", length(DIO$SatelliteTables$flows$Flowable), "unique flows, and",length(DIO$Indicators$meta$Name), "indicators."))

print(paste("The model IO Year is",DIO$specs$IOYear))

print(paste("The model is capable of performing dollar year adjustments to results for the following years:" ))
as.character(colnames(DIO[["Rho"]]))


```

## Validate the Model


Validate that commodity output can be recalculated (within 1%) with the model total requirements matrix (L) and demand vector (y) for US production
```{r}
econval <- compareOutputandLeontiefXDemand(DIO, tolerance = 0.01)
print(paste("Number of sectors passing:",econval$N_Pass))
print(paste("Number of sectors failing:",econval$N_Fail))
print(paste("Sectors failing:", paste(econval$Failure$rownames, collapse = ", ")))
```

Validate that flow totals by commodity (E_c) can be recalculated (within 1%) using the model satellite matrix (B), market shares matrix (V_n), total requirements matrix (L), and demand vector (y) for US production 
```{r}
modelval <- compareEandLCIResult(DIO, tolerance = 0.01)
print(paste("Number of flow totals by commodity passing:",modelval$N_Pass))
print(paste("Number of flow totals by commodity failing:",modelval$N_Fail))
#print(paste("Sectors failing:", paste(modelval$Failure$variable, collapse = ", ")))
```

Validate that commodity output can be recalculated (within 1%) with model total domestic requirements matrix (L_d) and model demand (y) for US production
```{r}
econval <- compareOutputandLeontiefXDemand(DIO,use_domestic=TRUE, tolerance = 0.01)
print(paste("Number of sectors passing:",econval$N_Pass))
print(paste("Number of sectors failing:",econval$N_Fail))
print(paste("Sectors failing:", paste(econval$Failure$rownames, collapse = ", ")))
```

Validate that commodity output are properly transformed to industry output via MarketShare
```{r}
q_x_val <- compareCommodityOutputXMarketShareandIndustryOutputwithCPITransformation(DIO, tolerance = 0.01)
print(paste("Number of flow totals by commodity passing:",q_x_val$N_Pass))
print(paste("Number of flow totals by commodity failing:",q_x_val$N_Fail))
print(paste("Sectors with flow totals failing:", paste(q_x_val$Failure$rownames, collapse = ", ")))
```

Validate that commodity output equals to domestic use plus production demand
```{r}
q_val <- compareCommodityOutputandDomesticUseplusProductionDemand(DIO, tolerance = 0.01)
print(paste("Number of flow totals by commodity passing:",q_val$N_Pass))
print(paste("Number of flow totals by commodity failing:",q_val$N_Fail))
print(paste("Sectors with flow totals failing:", paste(q_val$Failure$rownames, collapse = ", ")))
```